---
title: "Project"
author: "Jason Suwito"
date: "2025-04-10"
output: html_document
---

```{r, message=FALSE}
library(spatstat)
library(sp)
library(sf)

load('BC_Covariates.Rda')
BC_Cov <- DATA
BC_Cov <- na.omit(BC_Cov)
head(summary(BC_Cov))
bees <- read.csv('dataset.csv')
coordinates(bees) <- ~decimalLongitude + decimalLatitude
proj4string(bees) <- CRS("+proj=longlat +datum=WGS84")
aea_crs <- CRS("+proj=aea +lat_0=45 +lon_0=-126 +lat_1=50 +lat_2=58.5 +x_0=1000000 +y_0=0 +datum=NAD83 +units=m +no_defs")
bees_proj <- spTransform(bees, aea_crs)
bee_coords <- coordinates(bees_proj)
head(bees,2)
window_sf <- st_as_sf(DATA$Window)
win <- as.owin(window_sf) 
bee_ppp <- ppp(x = bee_coords[,1], y = bee_coords[,2], window = win)
```

```{r, cache=TRUE}
elev <- BC_Cov$Elevation
forest <- BC_Cov$Forest
hfi <- BC_Cov$HFI
distwater <- BC_Cov$Dist_Water

# Replace NA values with median for each covariate
elev[is.na(elev)] <- median(elev, na.rm = TRUE)
forest[is.na(forest)] <- median(forest, na.rm = TRUE)
hfi[is.na(hfi)] <- median(hfi, na.rm = TRUE)
distwater[is.na(distwater)] <- median(distwater, na.rm = TRUE)

rho_elev <- rhohat(bee_ppp, elev)
rho_forest <- rhohat(bee_ppp, forest)
rho_hfi <- rhohat(bee_ppp, hfi)
rho_distwater <- rhohat(bee_ppp, distwater)
```

```{r}
plot(rho_elev,
     main = "",
     xlab = "Elevation",
     xlim = c(0, max(elev)))
plot(rho_forest,
     main = "",
     xlab = "Forest Cover"
     )
plot(rho_hfi,
     main = "",
     xlab = "HFI"
     )
plot(rho_distwater,
     main = "",
     xlab = "Distance Water"
     )

# Check for collinearity
cor.im(elev, forest, hfi, distwater, use = "pairwise.complete.obs")
```

1.  Elevation

-   Strong non-linear effect.
-   High density of observations at low elevations, with a sharp drop-off.
-   Suggests an inverse or log-like smooth effect.
-   Justifies a non-linear smooth $f_1(\text{elev})$.

2.  Forest Cover

-   Non-linear relationship with peak density around 10–30% forest cover, declining afterward.
-   Suggests unimodal or possibly a polynomial-like smooth effect.
-   Justifies $f_2(\text{forest})$.

3.  Human Footprint Index (HFI)

-   Very sharp increase in density at high HFI values (\>0.8).
-   Indicates a steep non-linear response—possibly threshold-like.
-   Justifies $f_3(\text{hfi})$.

4.  Distance to Water

-   No consistent trend, but some localized peaks.
-   Appears to have weak or noisy association with density.
-   Still, some non-linear relationship is present.
-   Justifies using a smooth $f_4(\text{distwater})$, though effect may be small.

5.  Collinearity Check

-   Correlation coefficients are all \< 0.3 in magnitude.
-   No strong collinearity between variables.
-   All variables can be included without serious multicollinearity concerns.

Proposed Model Formula:


$$
\log(\lambda) = \beta_0 
+ \beta_1 \cdot \text{elev} 
+ \beta_2 \cdot \text{forest} + \beta_3 \cdot \text{forest}^2 
+ \beta_4 \cdot \text{hfi} + \beta_5 \cdot \text{hfi}^2 
+ \beta_6 \cdot \text{distwater}
$$

```{r, cache=TRUE}
covariates_list <- list(elev = elev, forest = forest, hfi = hfi, distwater = distwater)

covariates_list <- lapply(covariates_list, function(im) {
  im$v[is.na(im$v)] <- median(as.vector(im$v), na.rm = TRUE)
  return(im)
})

model_null <- ppm(bee_ppp ~ 1, covariates = covariates_list)
print("=== NULL MODEL: ===")
print(model_null)

model_proposed <- ppm(bee_ppp ~ 
                             elev + 
                             forest + I(forest^2) + 
                             hfi + I(hfi^2) + 
                             distwater,
                           covariates = covariates_list)
print("=== PROPOSED MODEL: ===")
print(model_proposed)

print("=== AIC NULL MODEL: ===")
AIC(model_null)
print("=== AIC PROPOSED MODEL: ===")
AIC(model_proposed)

print("=== ANOVA TEST: ===")
anova(model_null, model_proposed, test = "Chi")
```

```{r, cache=TRUE}
intensity_im <- predict(model_proposed, type = "trend", n = 200)
log_intensity_im <- log(intensity_im)

plot(log_intensity_im,
     se = FALSE,
     superimpose = FALSE,
     main = "Log Fitted Intensity")

plot(log_intensity_im,
     se = FALSE,
     superimpose = FALSE,
     main = "Log Fitted Intensity")

plot(bee_ppp,
     pch = 16,
     cex = 0.6,
     cols = "green",
     add = TRUE)

med_elev <- median(elev, na.rm = TRUE)
med_forest <- median(forest, na.rm = TRUE)
med_hfi <- median(hfi, na.rm = TRUE)
med_dist <- median(distwater, na.rm = TRUE)


elev_effect <- effectfun(model_proposed, "elev", 
                         forest = med_forest,
                         hfi = med_hfi,
                         distwater = med_dist,
                         se.fit = TRUE)

forest_effect <- effectfun(model_proposed, "forest", 
                           elev = med_elev,
                           hfi = med_hfi,
                           distwater = med_dist,
                           se.fit = TRUE)

hfi_effect <- effectfun(model_proposed, "hfi", 
                        elev = med_elev,
                        forest = med_forest,
                        distwater = med_dist,
                        se.fit = TRUE)

dist_effect <- effectfun(model_proposed, "distwater", 
                         elev = med_elev,
                         forest = med_forest,
                         hfi = med_hfi,
                         se.fit = TRUE)

plot(elev_effect,
     legend = FALSE,
     main = "Effect of Elevation\nat median forest, hfi, distance")
plot(forest_effect,
     legend = FALSE,
     main = "Effect of Forest\nat median elev, hfi, distance")
plot(hfi_effect,
     legend = FALSE,
     main = "Effect of HFI\nat median elev, forest, distance")
plot(dist_effect,
     legend = FALSE,
     main = "Effect of Distance to Water\nat median elev, forest, hfi")
```


```{r}
quadrat.test(model_proposed, nx = 5, ny = 5)
```

```{r}
res <- residuals(model_proposed)
print(res)
plot(res, main = "Model Residual", cols = "transparent")
```

```{r, cache=TRUE}
par_res_elev <- parres(model_proposed, covariate = "elev")
par_res_forest <- parres(model_proposed, covariate = "forest")
par_res_hfi <- parres(model_proposed, covariate = "hfi")
par_res_dist <- parres(model_proposed, covariate = "distwater")

plot(par_res_elev,
     legend = FALSE,
     lwd = 2,
     main = "Partial Residuals: Elevation",
     xlab = "Elevation (m)")

plot(par_res_forest,
     legend = FALSE,
     lwd = 2,
     main = "Partial Residuals: Forest Cover",
     xlab = "Forest Cover (%)")

plot(par_res_hfi,
     legend = FALSE,
     lwd = 2,
     main = "Partial Residuals: HFI",
     xlab = "Human Footprint Index")

plot(par_res_dist,
     legend = FALSE,
     lwd = 2,
     main = "Partial Residuals: Distance to Water",
     xlab = "Distance to Water (m)")
```
