---
title: "Project_lab2_Yuzhou"
author: "Yuzhou Wang"
date: "2025-04-16"
output: html_document
---

```{r}
library(spatstat.geom)
library(sf)
load('BC_Covariates.Rda')
ls()
head(summary(DATA))
bees <- read.csv('dataset.csv')
coordinates(bees) <- ~decimalLongitude + decimalLatitude
proj4string(bees) <- CRS("+proj=longlat +datum=WGS84")
aea_crs <- CRS("+proj=aea +lat_0=45 +lon_0=-126 +lat_1=50 +lat_2=58.5 +x_0=1000000 +y_0=0 +datum=NAD83 +units=m +no_defs")
bees_proj <- spTransform(bees, aea_crs)
bee_coords <- coordinates(bees_proj)
head(bees,2)
window_sf <- st_as_sf(DATA$Window)
win <- as.owin(window_sf) 

bee_ppp <- ppp(x = bee_coords[,1], y = bee_coords[,2], window = win)
table(inside.owin(bee_ppp$x, bee_ppp$y, win))  
plot(DATA$Elevation, main = "Elevation + Bee Observations")
plot(bee_ppp, add = TRUE, pch = 20, cex = 0.6)
```

```{r}

library(spatstat.geom)

bee_win_km <- rescale(Window(bee_ppp), s = 1000, unitname = "km")

npoints(bee_ppp) / area(bee_win_km)
 


```

In this section, we estimated the first-order intensity of bee observations across the study region. Initially, the intensity was calculated in units of points per square meter, resulting in an extremely small value (~1.17e-09), which is difficult to interpret. To make the results more intuitive, we rescaled the spatial window to units of square kilometers, yielding a new intensity estimate of ~0.00117 bees/km².  

This very low value suggests that bee observations are extremely sparse over the study area. While this could indicate a naturally low density of bees in the region, it is also possible that the bees are concentrated in specific locations, rather than being evenly distributed. This motivates further analysis of spatial inhomogeneity, using methods like quadrat counts, kernel density estimation, and second-order statistics such as Ripley’s K-function, which will be explored in the subsequent sections.  

```{r}
library(spatstat.explore)

Q <- quadratcount(bee_ppp, nx = 5, ny = 5)
plot(bee_ppp, pch = 20, cex = 0.5)
plot(Q, add = TRUE)
quadrat.test(bee_ppp, nx = 5, ny = 5)



```

```{r}
plot(intensity(Q, image = TRUE),
     main = "Beilschmiedia pendula intensity")


plot(bee_ppp,
     pch = 16,
     cex = 0.6,
     cols = "white",
     add = TRUE)

plot(bee_ppp,
     pch = 16,
     cex = 0.5,
     cols = "black",
     add = TRUE)

```


The quadrat-based intensity map of Beilschmiedia pendula shows clear spatial heterogeneity in the species’ distribution. The highest intensities are concentrated in the southern and southeastern parts of the study region, while most of the northern areas show near-zero intensity. This suggests that B. pendula is either ecologically constrained to certain environments or is under-sampled in other areas.


### Interpretation

In this analysis, I performed a quadrat test to evaluate whether the spatial distribution of bee observations in British Columbia follows a pattern of complete spatial randomness (CSR). The region was divided into a 5x5 grid (with 21 valid quadrats), and bee counts were tallied within each quadrat.  

The resulting map shows clear spatial heterogeneity in bee observations. Some quadrats, especially in the southern region, have very high counts (e.g., 322 and 66), while others—particularly in the north—have counts close to or exactly zero. This visual pattern already suggests a non-random distribution.  

The quadrat test produced the following results:  
Chi-squared = 7290.7, df = 20, p-value < 2.2e-16  
These results indicate a highly significant deviation from CSR. The extremely low p-value allows us to reject the null hypothesis that bee occurrences are randomly distributed across space.

### Conclusion

There is strong evidence that the bees are not distributed randomly, but rather show significant spatial clustering. This may reflect environmental, ecological, or human factors influencing bee presence in the region.


```{r}
library(spatstat.geom)
library(spatstat.explore)


lambda_u_hat <- density(bee_ppp) 

plot(lambda_u_hat,
     main = "Kernel estimate of bee_ppp intensity")

plot(bee_ppp,
     pch = 16,
     cex = 0.6,
     cols = "white",
     add = TRUE)

plot(bee_ppp,
     pch = 16,
     cex = 0.5,
     cols = "black",
     add = TRUE)


```



The kernel density estimate of bee_ppp shows that bee observations are strongly clustered in the southern part of the study area. The intensity surface highlights areas with the highest densities in yellow, suggesting these are potential hotspots for bee activity or observation effort. The presence of many zero- or low-intensity zones in the north may indicate ecological constraints, lower sampling effort, or habitat unsuitability.


```{r}

library(spatstat.geom)
library(spatstat.explore)

par(mfrow = c(1, 3), mar = rep(0.1, 4)) 

# 1. Diggle bandwidth
plot(density(bee_ppp, sigma = bw.diggle(bee_ppp)), 
     ribbon = FALSE,
     main = "Bandwidth: Diggle")

# 2. PPL bandwidth
plot(density(bee_ppp, sigma = bw.ppl(bee_ppp)), 
     ribbon = FALSE,
     main = "Bandwidth: PPL")

# 3. Adaptive KDE
lambda_u_hat_adaptive <- adaptive.density(bee_ppp, method = "kernel")
plot(lambda_u_hat_adaptive,
     main = "Adaptive kernel estimate")


```


The three kernel density maps illustrate different smoothing strategies for the bee point pattern (bee_ppp). The Diggle and PPL methods use fixed bandwidths, with PPL producing slightly sharper intensity contrasts. The adaptive kernel method adjusts the bandwidth based on local point density, resulting in enhanced resolution in dense areas while maintaining smoother estimates in sparse regions. However, all three maps show similar general patterns, with high-intensity hotspots concentrated in the southern region of the study area.

```{r}
library(spatstat.explore)

R <- bw.ppl(bee_ppp)

LR <- scanLRTS(bee_ppp, r = R)

plot(LR, main = "Local likelihood ratio test (bee_ppp)")

contour(LR, add = TRUE)
plot(bee_ppp, add = TRUE, pch = 16, cex = 0.4, cols = "black")

```

```{r}
pvals <- eval.im(pchisq(LR,
                        df = 1,
                        lower.tail = FALSE))
plot(pvals, main = "Local p-values")

```


The local likelihood ratio test (LRT) highlights strong spatial clustering of bee observations in the southern region. The corresponding p-value map confirms that these clusters are statistically significant, with p-values close to zero (blue areas). This suggests that the observed clustering is unlikely to have occurred under complete spatial randomness (CSR) and indicates potential environmental or ecological drivers influencing bee distribution.

```{r}
bee_elev <- DATA$Elevation
elev <- bee_elev

b <- quantile(elev, probs = (0:4)/4, type = 2)
Zcut <- cut(elev, breaks = b)
V <- tess(image = Zcut)

quadratcount(bee_ppp, tess = V)


rho <- rhohat(bee_ppp, elev)


plot(rho, main = "Intensity of bees vs elevation")

library(spatstat.explore)

env <- envelope(bee_ppp, Gest, nsim = 39)
plot(env, main = "CSR test via G-function envelope")




```

A spatial point pattern analysis was conducted to assess the relationship between bee intensity and elevation, as well as deviations from complete spatial randomness (CSR). The non-parametric estimate showed that bee intensity is strongly concentrated at lower elevations. A Monte Carlo envelope test using the G-function further revealed that the observed bee locations exhibit significant clustering, with the observed G-function lying well above the simulated CSR envelopes.



