---
title: "project_lab2_haoxiang"
author: "haoxiang_xu"
date: "2025-04-16"
output: html_document
---

```{r}
library(spatstat.geom)
library(spatstat)
library(sf)
```

```{r}
load('BC_Covariates.Rda')
ls()
head(summary(DATA))
bees <- read.csv('dataset.csv')
coordinates(bees) <- ~decimalLongitude + decimalLatitude
proj4string(bees) <- CRS("+proj=longlat +datum=WGS84")
aea_crs <- CRS("+proj=aea +lat_0=45 +lon_0=-126 +lat_1=50 +lat_2=58.5 +x_0=1000000 +y_0=0 +datum=NAD83 +units=m +no_defs")
bees_proj <- spTransform(bees, aea_crs)
bee_coords <- coordinates(bees_proj)
head(bees,2)
window_sf <- st_as_sf(DATA$Window)
win <- as.owin(window_sf)

bee_ppp <- ppp(x = bee_coords[,1], y = bee_coords[,2], window = win)
table(inside.owin(bee_ppp$x, bee_ppp$y, win))  
plot(DATA$Elevation, main = "Elevation + Bee Observations")
plot(bee_ppp, add = TRUE, pch = 20, cex = 0.6)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## Morisita’s Index plot

```{r}
rect_win <- as.owin(as.rectangle(win))
bee_ppp_rect <- ppp(x = bee_coords[,1], y = bee_coords[,2], window = rect_win)
miplot(bee_ppp_rect, main = "Morisita Index (Rectangular Window)", pch = 16, col = "#046C9A")
```

## Ripley’s K-function estimate

```{r}
xrange <- range(bee_coords[,1])
yrange <- range(bee_coords[,2])
win_rect <- owin(xrange = xrange, yrange = yrange)

unique_coords <- bee_coords[!duplicated(bee_coords), ]
bee_ppp_unique <- ppp(x = unique_coords[,1], y = unique_coords[,2], window = win_rect)
k_bee <- Kest(bee_ppp_unique)
plot(k_bee, main = "Ripley’s K-function (No Duplicates)", lwd = 2)
```

## K-function with bootstrapped confidence intervals (α=0.05)

```{r}
E_bee <- envelope(bee_ppp, Kest, correction = "border", rank = 1, nsim = 19, fix.n = TRUE)
plot(E_bee, main = "", lwd = 2)
```

## K-function with bootstrapped confidence intervals (α=0.01)

```{r}
E_bee_99 <- envelope(bee_ppp, Kest, correction = "border", rank = 1, nsim = 99, fix.n = TRUE)
plot(E_bee_99, main = "", lwd = 2)
```

## Estimate intensity for inhomogeneous K and PCF

```{r}
lambda_bee <- density(bee_ppp, sigma = bw.ppl(bee_ppp), positive = TRUE)
cat("Mean estimated intensity (λ):", mean(lambda_bee$v, na.rm = TRUE), "\n")
cat("Max estimated intensity (λ):", max(lambda_bee$v, na.rm = TRUE), "\n")
```

## Inhomogeneous K-function

```{r}
Kinhom_bee <- Kinhom(bee_ppp, lambda = lambda_bee)
plot(Kinhom_bee, theo ~ r, main = "", col = "grey70", lty = "dashed", lwd = 2)
plot(Kinhom_bee, border ~ r, col = "#046C9A", lwd = 2, add = T)
```

## Simulation envelope for inhomogeneous K-function

```{r}
E_bee_inhom <- envelope(bee_ppp, Kinhom, simulate = expression(rpoispp(lambda_bee)), correction = "border", rank = 1, nsim = 19, fix.n = TRUE)
par(mfrow = c(1, 2))
plot(E_bee_inhom, main = "", lwd = 2)
plot(E_bee_inhom, xlim = c(0, 50000), main = "", lwd = 2)
```
